\documentclass[12pt,letterpaper]{article}

% \usepackage[margin=0.9in, showframe]{geometry}
\usepackage[
    top=5mm,
    bottom=5mm,
    left=5mm,
    right=5mm,
    marginparwidth=0mm,
    marginparsep=0mm,
    headheight=15pt,
    centering,
    % showframe,
    includefoot,
    includehead
    ]{geometry}


\usepackage{setspace}
\usepackage{titlesec}
\usepackage{lipsum}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{xcolor}

\usepackage{blindtext}
\usepackage{multicol}
\usepackage{fancyhdr}

% add more margin between multicols columns
\setlength\columnsep{10pt} 

\vspace{-\baselineskip}

% ================== %
% format of sections
% ================== %
\titleformat{\section}
   {\normalfont\small\bfseries\MakeUppercase}{}{1em}{\underline}
\titlespacing{\section}{-1pc}{*4}{*1.5}

\titleformat{\subsection}
   {\normalfont\normalsize\bfseries}{}{1em}{}
\titlespacing{\subsection}{-0.7pc}{*4}{*1.5}

\titleformat{\subsubsection}
   {\small\bfseries\sc}{}{1em}{}
\titlespacing{\subsection}{-0.5pc}{*4}{*1.5}



% ================= %
% Headers & Footers
% ================= %
\pagestyle{fancy}
\fancyhf{}
\lhead{\bf STAT410 Notes}
\rhead{Gaetan Almela}
\rfoot{\thepage}
\renewcommand{\headrulewidth}{0pt}



% ================= %
% Aliases
% ================= %
\def\cov{\text{Cov}}
\def\var{\text{V}}
\def\ev{\text{E}}
\def\eps{\varepsilon{}}
\def\Dom{\text{Dom}}



% ================= %
% Commands
% ================= %
\newcommand{\Z}{{\mathbb Z}}
\newcommand{\Q}{{\mathbb Q}}
\newcommand{\R}{{\mathbb R}}
\newcommand{\D}{{\mathbb D}}
\newcommand{\N}{{\mathbb N}}

% Used for sigma algebras
% \newcommand{\B}{{\mathcal B}}

\newcommand{\btw}[1]{
    $\langle$ #1 $\rangle$
}

\newcommand{\powerset}{\raisebox{.15\baselineskip}{\Large\ensuremath{\wp}}}

\newcommand{\B}{\mathcal B}

% create a command to display examples in a box
\newcommand{\ex}[1]{
    \medskip \noindent
    \fcolorbox{lightgray}{white}{
        \parbox{0.8\linewidth}{
            {\bf Example} \\

            #1
        }
    } 
}

% create a command for TODOs
\newcommand{\TODO}{\color{red}\textbf{TODO}\color{black}}



\begin{document}
    \footnotesize

    \begin{multicols*}{2}
        \btw{Zoom recordings and notes are available through elms}

        \section{Probabilities}

        Goal is to understand how to calculate probabilities.

        \subsection{Experiments}

        An experiment is {\it a task that has 2 attributes}:
        \begin{enumerate}
            \item It is repeatable
            \item All outcomes are known and well defined.
        \end{enumerate}

        \ex {
            Pick out colored balls from a bag, you need to know what the different
            colors can be.

            Once an experiment has been designed, you will have the {\bf sample
            space $S$}: the set of all possible outcomes of the experiment.
        }

        \ex {
            Experiment: Toss a coin once.

            Sample Space: $\{H, T\}$. We assume that this is a two-sided coin.
        }

        \ex {
            Toss that same coin 4 times.

            Sample Space

            \[
                S = \begin{bmatrix}
                    HHHH & HHHT & HHTH & HHTT \\

                    HTHH & HTHT & HTTH & HTTT \\

                    THHH & THHT & THTH & THTT \\

                    TTHH & TTHT & TTTH & TTTT
                \end{bmatrix}
            \]

            There are 16 outcomes in this sample space. $|S| = 16$. Note that
            this space is 1 dimensional, I only formatted it as a matrix to save
            space.
        }

        \ex {
            Roll a 6-sided die once

            \[
                S = \begin{bmatrix} 
                    1 \\
                    2 \\
                    \vdots \\
                    5 \\
                    6
                \end{bmatrix}
            \]

            Roll a 6-sided die twice

            \[
                S = \begin{bmatrix} 
                    (1, 1) & \dots  & (1, 6) \\
                    \vdots & \ddots & \vdots \\
                    (6, 1) & \dots  & (6, 6) 
                \end{bmatrix}
            \]
        }

        \subsection{Events}
        An event is any subset $E$ of the sample space.

        \ex {
            \[
                E = \begin{bmatrix} 
                    (1, 1) \\
                    (2, 2) \\
                    \vdots \\
                    (5, 5) \\
                    (6, 6)
                \end{bmatrix}
            \]
        }

        \subsection{Probabilities of Events}

        There are $\powerset(S) = 2^{|S|}$ different subsets $E$ of $S$. This
        grows very quickly. In our case of rolling a die twice, the set of all
        subsets is $\powerset(S) = 2^{36}$.

        {\bf Note}: For a set $F$, $F^c = U - F$ is the complement of $F$
        where $U$ is the universe.

        \ex {
            For a set $F = \{x \leq 10\}$, $F^c = \{ x > 10 \}$
        }

        \subsection{Set Theory}

        Let $A$ and $B$ be sets. \\

        {\bf Complement}

        $A^c = \{ x \not\in A \}$ is the complement of $A$. \\

        {\bf Subsets}

        $A \subseteq B$ if and only if $x \in A \Rightarrow x \in B$ \\

        {\bf Union}

        $A \cup B = \{x | x \in A \text{ or } x \in B\}$ \\

        {\bf Intersections}

        $A \cap B = \{x | x \in A \text{ and } x \in B\}$ \\

        \btw{Goal: Quantify the probability of an event.}

        It is important to note that one should consider whether the events in
        the experiment are dependent or independent.

        \ex {
            Suppose the experiment is to roll a die and weight it. Then the
            sample space $S = \{ \text{weight of the die $\in \R$} \}$. This is
            true regardless of the number shown on the die, since these two
            values are {\it independent}.

            Note that $\R$ is uncountably infinite
        }

        If the sample space $S$ is finite, and $|S| = N$, any subset of $S$ is
        an event. There are $\powerset(S) = 2^N$ possible subsets.

        \ex {
            {\bf Toss a coin once.}

            Then $S = \{H, T\}$ so $|S| = 2$, then $|\powerset(S)| = 2^2 = 4$.
            \\

            {\bf Toss a coin twice.}

            Then $|S| = 4$, then $|\powerset(S)| = 2^4 = 16$.
            \\

            {\bf Roll a die once.}

            Then $|S| = 6$, then $|\powerset(S)| = 2^6 = 64$.
            \\

            {\bf Roll a die twice.}

            Then $|S| = 36$, then $|\powerset(S)| = 2^{36}$.
            \\

            {\bf Roll an $N$ sided die.}

            Then $|S| = N$, then $|\powerset(S)| = 2^N$.
            \\
        }

        {\bf Note} Event space is growing exponentially in the size of the
        sample space

        \section{$\sigma$ Algebras}

        \btw {{\bf Goal} Cut down the event space based on interesting/important
        events.}

        We will define probabilities only for events we are interested in.

        A collection of $\B$ sets is called a sigma algebra if
        \begin{itemize}
            \item $\emptyset \in \B$
            \item If $A \in \B$, then $A^c \in \B$ ($\B$ is closed under
                complements)
            \item If $A_1, A_2, A_3, ...$ is a countable collection of sets
                in $\B$, then $\bigcup_{i = 1} A_i \in \B$.
        \end{itemize}

        \ex {
            Toss the coin three times.

            $\powerset(S)$ is the largest possible sigma algebra with size
            $2^|S| = 2^8 = 256$.
        }

        \ex {
            Toss the coin three times, is the second toss a $H$?

            $|S| = 2^3 = 8$ but now we only care about 2 types of outcomes,
            either the middle toss is $H$ or $T$. \\

            Let $\{0\}$ be the set such that you toss a coin 3 times and second
            toss is $H \rightarrow \{HHH, HHT, THT, THH\}$

            Let $\{1\}$ be the set such that you toss a coin 3 times and second
            toss is $T \rightarrow \{TTT, TTH, HTT, HTH\}$ \\

            Let $B$ be the sigma algebra generated by $\{0\}$ and $\{1\}$, then

            \TODO {Format better}
            \begin{align*}
                B&= \{ \emptyset, \{0, 1\}, \{0\}, \{1\}\} \\
                 &= \{ \emptyset, S, \\
                 &      \{HHH, HHT, THH, THT\}, \\
                 &      \{TTT, TTH, HTT, HTH\} \\
                 &  \}
            \end{align*}

            For step 2, we just substituted $\{0\}$ and $\{1\}$.

            Note that the union of any two sets in $\B$ is also in $\B$. 
        }

        Note that the reason we don't just use $\powerset(S)$ all the time for
        probability is because the powerset gets huge so fast, even for
        computers. This is the problem that sigma algebras solve.

        This is useful to define a probability function. \\

        \subsection{Probability Function}
        Given a pair $(S, \B)$ where $S$ is the sample space and $\B$ is the sigma
        algebra. A {\bf probability function} $P: \B \rightarrow \R$ for this
        pair is a function on $\B$ satisfying 3 axioms

        \begin{itemize}
            \item $P(\emptyset) = 0$.
            \item $P(S) = 1$.
            \item $P(A) \geq 0$ for all $A \in \B$.
            \item If $A_1, A_2, A_3, ...$ is a collection of pairwise disjoint
                sets in $\B$, $P(\bigcup_{i} A_i) = \sum_{i = 1}^{\infty}
                P(A_i)$. Note that $S$ could be infinite, which means that since
                $\B \subseteq \powerset(S)$ each $A_i$ could be infinite.
        \end{itemize}

        \TODO Take notes on missing section

        \section{Conditional Probability}

        {\bf Recall} We want to assign probability functions, we came up with
        the definition of probability functions: $P(A | B)$ is the probability
        that $A$ happens given that $B$ has happened

        \[
            P(A | B) = \frac{P(A \cap B)}{P(B)}; P(B) > 0
        \]

        \subsection{Multiplication Principle}

        \[
            P(A \cap B) = P(A | B) \cdot P(B) = P(B | A) \cdot P(A)
        \]

        $P(A \cap B)$ is the probability of the interaction between $A$ and $B$
        times the probability of $B$.

        This leads us to the definition, $A$ and $B$ are independent events if
        
        \[
            P(A | B) = P(A) \cdot P(B) \Leftrightarrow P(A | B) = P(A)
            \Leftrightarrow P(B | A) = P(B)
        \]

        \subsection{Equally likely outcomes}

        Suppose $S$ is the sample space of an experiment, We assume that $N = |S|$
        is finite. We say that all outcomes are equally likely if $P(\{s\}) =
        \frac{1}{N}$

        \ex {
            Tossing a fair coin twice: $S = \{HH, HT, TH, TT\}$. Equally likely
            outcomes means that $P(\{HH\}) = P(\{HT\}) = \cdots = \frac{1}{4}$.
        }

        {\bf Note}: If $P(H) = \frac{1}{2}$ and the tosses are independent,
        then we have equally likely outcomes. \\

        If $N = |S|$ is finite, then all outcomes in $S$ are equally likely,
        then calculating probabilities of an event $E$ is the same as
        counting the simple outcomes in $E$.

        \[
            P(E) = \frac{|E|}{|S|}
        \]

        Is true if {\bf all outcomes are equally likely}.

        \ex {
            If $P(H) = 0.8$, $P(T) = 0.2$, then

            \begin{align*}
                P(HH)&= 0.8^2 = 0.64 \\
                P(HT)&= 0.8 \cdot 0.2 = 0.16 \\
                P(TT)&= 0.2^2 = 0.04 \\
            \end{align*}

            This does not yield equally likely outcomes, as expected.
        }

        If $S$ has equally likely outcomes $\Leftrightarrow$ Calculating
        probabilities is the same as counting.

        \btw{{\bf Goal}: develop counting techniques.}

        \section{Fundamental Thm. of Counting}

        If a task $T$ can be broken down into a sequence of sub-tasks $T_1, T_2,
        ..., T_k$ such that $T: T_1 \rightarrow T_2 \rightarrow T_3 \rightarrow
        \cdots \rightarrow T_k$.

        \btw{This notation isn't function, it's doing task $T_1$, then $T_2$, then ...}

        How many ways can this be done?

        \[
            \text{\# ways to do task } T = n_1 \cdot n_2 \cdot \cdots \cdot n_k
        \]

        Where $n_i$ is the number of ways that task $T_i$ can be done.

        This can be proven with induction.

        \subsection{Counting Techniques}

        There are four possible meanings to ``From $n$, pick $k$" \\

        {\bf Allow Replacement, Order matters.}

        \[
            n^k
        \]

        If you allow replacement and count any order, you have $n$
        options for each of your $k$ picks and {\it all permutations are
        distinct}. \\

        {\bf Allow Replacement, Order doesn't matter.}

        \[
            \frac{n^k}{k!}
        \]

        This is the same as above except order doesn't matter, which
        means that ``1, 2" is the same as ``2, 1". How many ways are
        there to permute $k$ things? $k!$ ways, so we divide the first
        result by $k!$ to remove all the extra counts. \\

        {\bf Don't allow replacement, Order matters.}

        \[
            _{n}P_k = \frac{n!}{(n - k)!}
        \]

        By not allowing replacement, it means that (for your first pick)
        you have $n$ choices, for the next, you have $n - 1$, the next
        has $n - 2$, and so on... However you're only picking $k$
        things, so this is why we ``stop this factorial early", after
        $k$ iterations. \\

        {\bf Don't allow replacement, Order doesn't matter.}

        \[
            _{n}C_k = \frac{n!}{k!(n - k)!}
        \]

        This is the same as above but now, we again add the condition
        that ``1, 2" is the same as ``2, 1", so we divide by $k!$ to
        remove all extra counts. \\

        This is useful for the Fundamental Theorem of Counting. Note that
        you should still understand where these formulas come from.

        %
        % Note about homework, 6 and 7 require you to count: 
        %
        % P(exactly one cell empty) = (# outcomes with one cell empty) / total #
        % of outcomes
        %
        % T1. choose the cell with 2 balls
        % T2. choose the cells with 1 ball each
        % T3. choose the cell that has no ball
        %

        %
        % hint for question 3
        % every dart hits the dartboard
        %
        % P(area of the region /total area)

    \end{multicols*}
\end{document}

